Typesafe Coordinate Systems in High-Throughput Sequencing Applications

### 2021-08-31 manuscript/main.tex
Footnotes require hyperref loaded last
manuscript wording update (Nic Wilson)

\documentclass[12pt]{article}
%not needed with LuaTex \usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage[hyphens]{url}   % required to make url's linebreak at -
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    urlcolor=blue,
    citecolor=magenta,
}

@@

@@

Instead, consider strongly and statically typed languages with generic and polymorphic features. Define a type class \texttt{Interval} of which there are four types, corresponding to ZBHO, ZBC, OBHO, and OBC coordinate systems, and for which constituent types have a \textit{to} (or conversely, \textit{from}) type conversion function. Note that the types may be freely interconverted, as there exists a bijective mapping between any two coordinate systems (Figure \ref{fig:transitiongraph}). While these types could all be defined individually, one might instead leverage polymorphic features of the type system, especially in systems where a type can be parameterized on a constant value. In C++, D, or similar languages this could be done by templating a single \texttt{Interval} class or struct on some enums defining basis and openness:

@@

Both software engineering and bioinformatics are relatively newer disciplines among the applied sciences. In the same way that other engineering fields may apply and benefit from fundamental advances in mathematics and physics, the integrity, safety, and performance of software systems can be improved by applying lessons from computer science. These principles have also made their way to bioinformatics and computational biology, where authors have recognized the Rust language (\href{http://www.rust-lang.org}{http://www.rust-lang.org}) for its memory safety attributes and its potential contributions to safer and more correct bioinformatics tools.\cite{koster_rust-bio_2016,perkel_why_2020}  Here, we demonstrate that leveraging the type system and compile-time guarantees of strongly typed languages like D, Rust, and others can reduce the error surface of software in the computational biology and bioinformatics space. These principles could also be extended to more loosely typed language such as Python, albeit with much weaker guarantees and a runtime overhead.

The reference implementation of Typesafe Coordinates as part of \textit{dhtslib}\cite{gregory_dhtslib_nodate} --- to our knowledge the first such polymorphic implementation of this concept in any bioinformatics tooling --- is written in D, a statically-typed, compiled language. We found D's compile-time introspection and metaprogramming features ideal for construction of such a system. We also implemented Typesafe Coordinates in other languages, including Rust, Python, and OCaml, albeit with language-specific limitations in each of these.

@@

>>>>

\documentclass[12pt]{article}
%not needed with LuaTex \usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage[hyphens]{url}   % required to make url's linebreak at -

@@

% Rule of thumb: load `hyperref` last.
% Why I love *eyeroll* LaTeX: Bizzare interaction of `setspace` and `hyperref`,
% both of which modify the \footnote command:
% https://tex.stackexchange.com/questions/188273/using-setspace-package-breaks-footnotes-link
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    urlcolor=blue,
    citecolor=magenta,
}

@@

Instead, consider strongly and statically typed languages with generic and polymorphic features. Define a type class \texttt{Interval} of which there are four types, corresponding to ZBHO, ZBC, OBHO, and OBC coordinate systems, and for which there exists a function, \textit{to}, (or conversely, \textit{from}) that maps between constituent types. Note that the types may be freely interconverted, as there exists a bijective mapping between any two coordinate systems (Figure \ref{fig:transitiongraph}). While these types could all be defined individually, one might instead leverage polymorphic features of the type system, especially in systems where a type can be parameterized on a constant value. In C++, D, or similar languages this could be done by templating a single \texttt{Interval} class or struct on some enums defining basis and openness:

@@

Both software engineering and bioinformatics are relatively newer disciplines among the applied sciences. In the same way that other engineering fields may apply and benefit from fundamental advances in mathematics and physics, the integrity, safety, and performance of software systems can be improved by applying lessons from computer science. These principles have also made their way to bioinformatics and computational biology, where authors have recognized the Rust language (\url{https://www.rust-lang.org}) for its memory safety attributes and its potential contributions to safer and more correct bioinformatics tools.\cite{koster_rust-bio_2016,perkel_why_2020}  Here, we demonstrate that leveraging the type system and compile-time guarantees of strongly typed languages like D, Rust, and others can reduce the error surface of software in the computational biology and bioinformatics space. These principles could also be extended to more loosely typed language such as Python, albeit with much weaker guarantees and a runtime overhead.

The reference implementation of Typesafe Coordinates as part of \textit{dhtslib}\cite{gregory_dhtslib_nodate} --- to our knowledge the first such polymorphic implementation of this concept in any bioinformatics tooling --- is written in D (\url{https://dlang.org/}), a statically-typed, compiled language. We found D's compile-time introspection and metaprogramming features ideal for construction of such a system. We also implemented Typesafe Coordinates in other languages, including Rust, Python, and OCaml, albeit with language-specific limitations in each of these.

@@

\section*{Acknowledgements} %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The authors wish to thank Nicholas Wilson for critically reading the draft manuscript.
